``Interestingness'' in Mathematical Theory Exploration

### 2019-10-18 PhDThesis/litreview_machinelearning.tex
Bunch more changes to thesis
Headingy things
More literview

\subsection{Machine Learning Applications in Mathematics, Logic and Programming Language Theory}

Machine learning techniques, used for finding statistical patterns and numerical
correlations in arbitrary data, are characterised by their statistical pattern-finding

AutoTuning

DeepTune analyses raw program text to predict the best options for an optimising
compiler. FIXME

Fuzzing AFL DeepSmith

>>>>

\section{Machine Learning Applications in Mathematics, Logic and Programming
  Language Theory}

Machine Learning (ML) applies statistical techniques to find patterns in data.
ML algorithms are numerical in nature, and often rely on assumptions such as
continuity and differentiability. In contrast, the worlds of mathematical
formulae, formal logic and computer programs are usually discrete and symbolic.
This makes application of ML to these domains more difficult than, say, image
processing.

One application of ML in these domains is to analyse real-world measurements
parameterised by some discrete entity, such as time or power used to execute
computer programs. Since flipping a single bit can cause arbitrarily large
changes in program behaviour (if it is the scrutinee of a conditional, for
example), such programs must be very similar for statistical patterns to emerge
in their executions. In the case of \emph{auto-tuning}, many programs variants
are measured, which all perform the same task in slightly different ways (e.g.
using different compiler optimisations, or dividing work differently between
hardware components); in this case the goal is to learn a predictive model of
which variant performs best for a given input~\cite{ganapathi2009case}.

Whilst performance measurements are simple numeric values, the symbolic values
need some conversion to become amenable to learning. This is usually performed
by a \emph{feature extraction} step, which takes various measurements (e.g.
counting occurrences of some element, assigning indices to different constants,
etc.) and packs them into a vector for subsequent analysis. More recently, so
called \emph{deep learning} systems have out performed those with manually
chosen features. Deep learning doesn't actually avoid feature extraction, but
its use of many ``layers'' of processing between the input and output makes the
selection of those features less important. For example, the feature extraction
in~\cite{cummins2017synthesizing} represents computer programs (after some
normalisation to a consistent layout and naming scheme) as a sequence of
individual characters, with each ASCII character replaced by a different number.
Yet even this crude method is enough for the deep learning system that consumed
these sequences to infer structure relevant for deciding whether it was better
suited to running on a CPU or GPU. The \textsc{DeepTune} system encodes slightly
more domain knowledge by tokenising language keywords, but is otherwise
similar~\cite{cummins2017end}.

\emph{Generative models} are another useful application of ML techniques to
symbolic systems. The \textsc{DeepSmith} system analyses a corpus of source code
to build a model of the programming language, which can then generate realistic
fragments of code, suitable for fuzz-testing compilers for that
language~\cite{cummins2017deepsmith}.

### 2019-10-18 PhDThesis/litreview_premiseselection.tex
Headingy things
More literview

\subsection{Relevance Filtering}
\label{sec:relevance}

\emph{Relevance filtering} takes a specified ``target'' (such as the current
goal of a theorem prover), a large list of inputs (e.g. existing lemmas from a
theory) and produces a sub-set of those inputs which are predicted to be useful
or relevant to the target. The example of choosing a sub-set of lemmas to help
prove a theorem is known as \emph{premise
  selection}~\cite{kuhlwein2012overview}.

>>>>

\section{Relevance Filtering}
\label{sec:relevance}

In the context of theorem proving, \emph{relevance filtering} takes a specified
``target'' (such as the current goal of a theorem prover), a large list of
inputs (e.g. existing lemmas from a theory) and produces a sub-set of those
inputs which are predicted to be useful or relevant to the target. The example
of choosing a sub-set of lemmas to help prove a theorem is known as
\emph{premise selection}~\cite{kuhlwein2012overview}.

### 2019-10-18 PhDThesis/litreview_staticanalysis.tex
More literview

Software for analysing and improving code can be roughly divided into two
categories. Some contribute to the running of a program, such as (optimising)
compilers, linkers and interpreters. These must produce correct, predictable,
desirable results in a timely manner. For example, compiler optimisations should
not alter program semantics, should provide monotonic improvements in the
majority of cases (i.e. they should not be \emph{pessimisations}) and conditions
under which they apply should be predictable enough to prevent spurious
performance regressions when making seemingly innocuous
refactorings~\cite{robison2001impact}. Optimisations in ahead-of-time (AOT)
compilers must be fast enough to run during every build, and should have a large
enough impact on the generated code's performance that they
``pay for themselves''~\cite{Franz1994}; for just-in-time (JIT) compilers such
ideals become hard requirements, since any runtime transformation which doesn't
pay for itself is, by definition, a pessimisation.

@@

To find project-specific improvements we need some description or specification
of how our program \emph{should} behave, to judge proposed improvements against.
Few projects have a complete formal specification, due to the prohibitive cost
of creating, validating, verifying and maintaining such artefacts. Partial
specifications, in the form of automated test suites, are more common. These
have been used to automatically patch errors in code (i.e. altering code until
its test suite passes), using methods from evolutionary
computation~\cite{Forrest.Nguyen.Weimer.ea:2009,weimer2009automatically,Schulte.Forrest.Weimer:2010}.

Optimisations treat \emph{the code itself} as a specification: formally
describing the desired behaviour, but not prescribing the exact steps that
should be taken to implement it. This allows different implementations to be
used as long as the result is equivalent\footnote{Optimisation assumes that
  extra efficiency is always desirable. In some areas that is not the case, such
  as security-sensitive code, where speed ups may expose sensitive information
  to timings attacks~\cite{kocher1996timing}.}. Taking this to the extreme gives
rise to \emph{superoptimisation}~\cite{massalin1987superoptimizer}, which
performs an open-ended search through the space of all programs to find those
equivalent to the given code, which are also optimal according to some quality
measure or cost model. Superoptimisers, such a
s \textsc{Souper}~\cite{sasnauskas2017souper} and \textsc{Aha!} (``A Hacker's
Assistant'', developed to accompany \emph{Hacker's
  Delight}~\cite{warren2013hacker}), can take an arbitrarily long time to find
improvements, but their optimisation power is limited only by the undecidability
of the source language (optimisations which cannot be proven equivalent are
necessarily ignored, like the uncomputable improvements demonstrated by
Blum~\cite{blum1967machine}).

@@

Superoptimisers also need some method to check whether a synthesised program
implements the required specification, such as a test suite and/or a theorem
prover. Proving correctness of machine code fragments is helped by the uniform,
finite nature of their domain (a fixed set of instructions, operating on a fixed
set of fixed-size registers). Higher level languages like Haskell are more
difficult to reason about and measure, since their behaviour and efficiency may
depend on unknown runtime parameters. In particular, optimising higher-order
functions is difficult in isolation, since many performance tradeoffs cannot be
decided without more knowledge of the functions involved. For example, a cache
of previous return values (known as a \emph{memo table}) can avoid re-computing
repeated calls to a function; but whether this actually saves time, or is worth
the memory penalty, depends on how long that function takes to calculate; and
how often it will be called with same arguments.

In addition to this uncertainty, each function's input domain may be infinite
(such as the natural numbers), and programs may use abstract interfaces and
datatypes with a variety of possible implementations (such as lists or key/value
mappings). All these factors makes effective testing more difficult, and proofs
potentially undecidable.

@@

Rather than optimising individual units of code in isolation, we can instead
work ``top down'' and perform \emph{whole program
  optimisation}~\cite{weeks2006whole,siskind1999flow}. In particular, for every
function we may want to optimise, we know \emph{all} call sites of that function
within the program, and hence avoid the need for a general solution.
Statically-known values, including arguments to (higher-order) functions, can be
propagated through the code using techniques like
\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}, and constant
expressions can be ahead of time; however this can dramatically increase code
size, as each function may be duplicated into multiple specialised instances.

Related techniques, like \emph{symbolic execution} (a form of \emph{abstract
  interpretation}), allow information to be statically deduced by tracing the
behaviour of code using an alternative execution model, where unknown runtime
data are represented as opaque metavariables. Property- and model-checkers are
similar, except they work dynamically by performing multiple runs, instantiating
metavariables with exhaustive or synthetic data (\quickcheck{} does this
randomly, \smallcheck{} enumerates values, etc.).

The most common code-improvement technique performed \emph{during execution} is
Just In Time (JIT) compilation~\cite{aycock2003brief}, which allows an
interpreter to compile sections of code into a faster form (usually machine
code). Since this compilation takes time, it is usually reserved for sections
which are executed often enough to exceed some predefined threshold (e.g. 100).
By focusing on such ``hot paths'', the up-front costs of compiling everything
ahead of time are avoided; although compiled forms are kept only in memory,
getting recompiled as needed on each program run. Compiling at runtime also
allows specialisation with dynamically-known information; for example, inlining
specific implementations of a procedure rather than performing dynamic dispatch
(this is the problem higher-order functions discussed above). To avoid altering
semantics, such dynamic specialisations must be accompanied by a conservative
validity check (e.g. that the given arguments should use the procedure
implementation we have compiled-in), falling back to the interpreter upon
failure.

@@

\begin{itemize}
\item Process A searches for proofs (in a pre-specified formal system) that some
  program $p\prime$ is both semantically equivalent to $p$ and has a running
  time bounded by some other program $t$, i.e. for all inputs $i$, the
  time taken to compute $p\prime(i)$ is bounded by $t(i)$. A list is built up of
  the $(p\prime, t)$ pairs found so far.
\item Process B computes $t(x)$ for all of the bounds $t$ found so far by A.
  Computations are interleaved via a schedule similar to Levin's universal
  search algorithm~\cite{levin1984randomness,ming}, to prevent getting stuck on
  non-halting programs. Whenever a calculation of some $t(x)$ finishes, it is
  compared to $t_{fast}$: if smaller, $t_{fast}$ is set to $t(x)$, and $p_{fast}$ is
  set to $p\prime$ (the program paired with this $t$).
\end{itemize}

>>>>

One obvious application of software properties discovered through Theory
Exploration is to improve that software, for example by \emph{refactoring} (e.g.
simplifying definitions) or \emph{optimising} (e.g. replacing a slow calculation
with an equivalent fast one). Existing tools for analysing and improving code
can be roughly divided into two categories. Some contribute to the running of a
program, such as (optimising) compilers, linkers and
interpreters\footnote{Optimisation has traditionally been the domain of
  compilers, but the recent popularity of \emph{link time optimisation} and
  \emph{just-in-time compilation} have brought the latter tools firmly under the
  code improvement umbrella.}. These must produce correct, predictable,
desirable results in a timely manner. For example, compiler optimisations should
not alter program semantics, should provide monotonic improvements in the
majority of cases (i.e. they should not be \emph{pessimisations}) and conditions
under which they apply should be predictable enough to prevent spurious
performance regressions when making seemingly innocuous
refactorings~\cite{robison2001impact}. Optimisations in ahead-of-time (AOT)
compilers/linkers must be fast enough to run during every build, and should have
a large enough impact on the generated code's performance that they
``pay for themselves''~\cite{Franz1994}; for just-in-time (JIT) compilers such
ideals become hard requirements, since any runtime transformation which doesn't
pay for itself is, by definition, a pessimisation.

@@

To find project-specific improvements we need some description or specification
of how our program \emph{should} behave, to judge proposed improvements against.
Few projects have a complete formal specification, due to the prohibitive cost
of creating, validating, verifying and maintaining such artefacts. Partial
specifications, in the form of automated test suites, are more common. These
have been used to automatically patch errors in code (i.e. altering code until
its test suite passes), using methods from evolutionary
computation~\cite{Forrest.Nguyen.Weimer.ea:2009, weimer2009automatically,
  Schulte.Forrest.Weimer:2010}. Such methods are necessarily standalone, since
making such arbitrary changes to a program's behaviour without a full
specification or user supervision has the potential to break the program in
untested ways.

Optimisations treat \emph{the code itself} as a specification: formally
describing the desired behaviour, but not prescribing the exact steps that
should be taken to implement it. This allows different implementations to be
used as long as the result is equivalent\footnote{Optimisation assumes that
  extra efficiency is always desirable. In some areas that is not the case, such
  as security-sensitive code, where speed ups may expose sensitive information
  to timings attacks~\cite{kocher1996timing}.}. Taking this to the extreme gives
rise to \emph{superoptimisation}~\cite{massalin1987superoptimizer}, which
performs an open-ended search through the space of all programs to find those
equivalent to the given code, which are also optimal according to some quality
measure or cost model. Superoptimisers, such as
\textsc{Souper}~\cite{sasnauskas2017souper} and \textsc{Aha!} (``A Hacker's
Assistant'', developed to accompany \emph{Hacker's
  Delight}~\cite{warren2013hacker}), can take an arbitrarily long time to find
improvements, but their optimisation power is limited only by the undecidability
of the source language (optimisations which cannot be proven equivalent are
necessarily ignored, like the uncomputable speedups demonstrated by
Blum~\cite{blum1967machine}).

@@

Superoptimisers also need some method to check whether a synthesised program
is equivalent to that given: either partially, like a test suite, or robustly
via a theorem prover. Proving correctness of machine code fragments is helped by
the uniform, finite nature of their domain (a fixed set of instructions,
operating on a fixed set of fixed-size registers). Higher level languages like
Haskell are more difficult to reason about and measure, since their behaviour
and efficiency may depend on unknown runtime parameters. In particular,
optimising higher-order functions is difficult in isolation, since many
performance tradeoffs cannot be decided without more knowledge of the functions
involved. For example, introducing a cache of previous return values (known as a
\emph{memo table}) can avoid re-computing repeated calls to a function; but
whether this actually saves time, or is worth the memory penalty, depends on how
long that function takes to calculate and how often it will be called with same
arguments.

In addition to this uncertainty, high level languages allow the input domain of
a function to be infinite (such as the natural numbers or lists), and programs
may use abstract interfaces and datatypes which either hide characteristics
required for establishing equivalence or cost, or may permit \emph{multiple}
implementations with varying characteristics. All these factors make effective
testing more difficult, and proofs potentially undecidable.

@@

Rather than optimising individual units of code in isolation, we can instead
work ``top down'' and perform \emph{whole program
  optimisation}~\cite{weeks2006whole,siskind1999flow}. In particular, for every
function we may want to optimise, we know \emph{all} call sites of that function
within the program, and hence avoid the need for a general solution.
Statically-known values, including arguments to (higher-order) functions, can be
propagated through the code using techniques like
\emph{supercompilation}~\cite{Turchin:1986:CS:5956.5957}, and constant
expressions can be computed ahead of time; however this can dramatically
increase code size, as each function may be duplicated into multiple specialised
instances. Link-time optimisation permits a limited form of inter-procedural
optimisation (e.g. inlining function definitions which were abstract black-boxes
during compilation), but generally avoids the more invasive optimisations found
in compilers.

Other methods to determine function usage and behaviour include \emph{symbolic
  execution} (a form of \emph{abstract
  interpretation}), which allow information to be statically deduced by tracing
the behaviour of code using an alternative execution model, where unknown
runtime data are represented as opaque metavariables. Property- and
model-checkers are similar, except they work dynamically by performing multiple
runs, instantiating metavariables with exhaustive or synthetic data
(\quickcheck{} does this randomly, \smallcheck{} enumerates values, etc.).

The most common code-improvement technique performed \emph{during execution} is
just-in-time (JIT) compilation~\cite{aycock2003brief}, which allows an
interpreter to compile sections of code into a faster form (usually machine
code). Since this compilation takes time, it is usually reserved for sections
which are executed often enough to exceed some predefined threshold (e.g. 100).
By focusing on such ``hot paths'', the up-front costs of compiling everything
ahead of time are avoided; although compiled forms are kept only in memory,
getting recompiled as needed on each program run. Compiling at runtime also
allows specialisation with dynamically-known information; for example, inlining
specific implementations of a procedure rather than performing dynamic dispatch
(this is the problem higher-order functions discussed above). To avoid altering
semantics, such dynamic specialisations must be accompanied by a conservative
validity check (e.g. that the given arguments should use the procedure
implementation we have compiled-in), falling back to the interpreter upon
failure.

@@

\begin{itemize}
\item Process A searches for proofs (in a pre-specified formal system) that some
  program $p'$ is both semantically equivalent to $p$ and has a running
  time bounded by some other program $t$, i.e. for all inputs $i$, the
  time taken to compute $p'(i)$ is bounded by $t(i)$. A list is built up of
  the $(p', t)$ pairs found so far.
\item Process B computes $t(x)$ for all of the bounds $t$ found so far by A.
  Computations are interleaved via a schedule similar to Levin's universal
  search algorithm~\cite{levin1984randomness,ming}, to prevent getting stuck on
  non-halting programs. Whenever a calculation of some $t(x)$ finishes, it is
  compared to $t_{fast}$: if smaller, $t_{fast}$ is set to $t(x)$, and $p_{fast}$ is
  set to $p'$ (the program paired with this $t$).
\end{itemize}
